### 2015-05-14 whoop.tex

To make our analysis scalable without sacrificing precision, and to cater for recursion, we use the Houdini~\cite{flanagan2001houdini} invariant inference algorithm to automatically compute summaries (pre- and post-conditions and loop invariants) from a pool of \emph{candidate} invariants. The Boogie verifier then automatically checks whether each procedure meets its summary, abstracting each called procedure with its summary in the process. Boogie also checks whether loops respect their invariants and uses the invariants to over-approximate the effects of loops.

The candidate invariants are based on the watchdog constants that \whoop instruments for watchdog race-checking. Our candidate generation rules are purely heuristic. The key idea is to automatically generate candidates for each memory region in the granularity of watched accesses. For example, if entry point $T$ has a procedure $f$ that only writes (and not reads) the offset $z$ of memory region $\mathit{MR}$, which is associated with the write set $\hasbeenwritten{\mathit{MR}, i}$ and read set $\hasbeenread{\mathit{MR}, i}$, then when instrumenting $T$ with symbol $i$ (using the notation from \S\ref{sec:symbolicpairwise}), we might guess $\mathit{watched}_{\mathit{MR}} = z \Longrightarrow z \in \hasbeenwritten{\mathit{MR}, i}$ and $\mathit{watched}_{\mathit{MR}} = z \Longrightarrow z \notin \hasbeenread{\mathit{MR}, i}$, as postconditions for $\instrument{f}{i}$. Likewise, if $f$ releases mutex $m$, then we might guess $m \notin \currentlockset{i}$ as a postcondition.

Using watchdog-based candidates reduces the false positives from reporting errors if only e.g.\ a field in memory region $\mathit{MR}$ is racing, but the rest of the fields in $\mathit{MR}$ are not racing. To generate the candidate invariants, \whoop performs an inter-procedural static analysis on the call-graph (in Boogie IVL code) of each entry point, identifying all possible accesses to shared memory locations. To be sound, if \whoop cannot infer information for a specific access, it generates a coarse-grained candidate for the corresponding memory region, e.g.\  $z \in \hasbeenwritten{\mathit{MR}, i}$ and $z \notin \hasbeenread{\mathit{MR}, i}$ for the above example.

We emphasize that the candidate invariants generated by \whoop are just that: candidates. The tool is free to speculatively generate candidates that later turn out to be incorrect: these are simply discarded by Houdini. A consequence is that incorrect or unintended candidates generated due to bugs in \whoop cannot compromise the soundness of verification.

>>>>

To make our analysis scalable without sacrificing precision, and to cater for recursion, we use the Houdini~\cite{flanagan2001houdini} invariant inference algorithm to automatically compute summaries (pre- and post-conditions and loop invariants) from a pool of \emph{candidate} invariants.  Given a pool of candidate invariants, Houdini repeatedly tries to verify each procedure with respect to its current candidate summary.  If verification fails due to an incorrect candidate, this candidate is discarded.  The process repeats until a fixpoint is reached, whereby the largest conjunction of the candidate invariants that can be proven to hold reamins.  The remaining candidates, which are indeed invariants, can then be assumed during further analysis of the program.

The Houdini algorithm does not generate the initial pool of candidate invariants; these are given to the algorithm as a starting point.  \whoop generates an initial pool of candidates using a set of heuristics.  The idea is to automatically generate likely program invariants based on syntactic patterns extracted from an inter-procedureal pass over the code for an entry point.  We give two examples; for clarity we use notation from the simple shared variable concurrent programming model of \S\ref{sec:symbolicpairwise}.  If we observe syntactically that procedure $f$ of entry point $T$ may write to, but does not read from, shared variable $s$.  Then when instrumenting $T$ with symbol $i$, we guess $s \in \hasbeenwritten{i}$ and $s \notin\hasbeenread{i}$ as postconditions for $\instrument{f}{i}$.  These guesses may be incorrect, for instance if the potential write to $s$ turns out to be in dead code, or if a read from $s$ has already been issued on entry to $\instrument{f}{i}$.  Similarly, if syntactic analysis indicates that $f$ may unlock mutex $m$, we guess $m \notin \currentlockset{i}$ as a postcondition for $\instrument{f}{i}$; this guess may be wrong, e.g.\ if the unlock operation is not reachable or if a subsequent lock operation acquires the mutex again.  We stress that guessing of incorrect candidate invariants does not compromise the soundness of verification: \whoop is free to speculatively generate candidates that are later deemed to be incorrect, and thus discarded, by Houdini.  The balance we try to strike is to have \whoop generate sufficient candidates to enable precise lockset analysis, without generating so many candidates that the speed of the Houdini algorithm is prohibitively slow.

In practice, the candidates are generated at the Boogie level, with respect to SMACK-generated memory regions and referencing the auxiliary variables introduced by our watchdog race instrumentation.
